GEE models

### 2017-10-15 doc/manuscript.Rmd
mL -> l

Methods: In the PROMISE longitudinal cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years. Outcome variables were calculated using the OGTT values. Insulin sensitivity was assessed using HOMA2-%S and the Matsuda index, while beta cell function was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were used, adjusting for time, waist, sex, ethnicity, baseline age, alanine aminotransferase, and physical activity. NEFA were analysed as both concentrations (nmol/mL) and proportions (mol%) of the total fraction.

@@

tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

@@

The primary outcome variables for this analysis were 
HOMA2-%S, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from further GEE analyses. There were `r meds_use[1]` and
`r meds_use[2]` participants who used lipid and cholesterol modifying medications
at the baseline and 6-year visit, respectively. Medication use did not influence
NEFA composition, the outcomes, or the GEE models (data not shown).

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta cell
function over 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty acids as a
proportion, an observation that was further confirmed by multivariate cluster
techniques, which found no ability of these individual NEFA to predict beta cell
function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
(20:5n-3), had a positive association with insulin sensitivity. In general,
these observations suggest that the absolute size of the total NEFA fraction,
rather than its specific composition (aside from EPA), likely influences the 
pathogenesis of diabetes, at least within a population at-risk for diabetes.

>>>>

Methods: In the PROMISE longitudinal cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years. Outcome variables were calculated using the OGTT values. Insulin sensitivity was assessed using HOMA2-%S and the Matsuda index, while beta cell function was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were used, adjusting for time, waist, sex, ethnicity, baseline age, alanine aminotransferase, and physical activity. NEFA were analysed as both concentrations (nmol/ml) and proportions (mol%) of the total fraction.

@@

tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
fig_dist <- fig('nefa', 'Concentrations (nmol/ml) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/ml) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/ml) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

@@

The primary outcome variables for this analysis were 
HOMA2-%S, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA as mole percent (mol%) of the total fraction and as
concentration (nmol/ml). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from further GEE analyses. There were `r meds_use[1]` and
`r meds_use[2]` participants who used lipid and cholesterol modifying medications
at the baseline and 6-year visit, respectively. Medication use did not influence
NEFA composition, the outcomes, or the GEE models (data not shown).

@@

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta cell
function over 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/ml) modeling, no associations were seen for these fatty acids as a
proportion, an observation that was further confirmed by multivariate cluster
techniques, which found no ability of these individual NEFA to predict beta cell
function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
(20:5n-3), had a positive association with insulin sensitivity. In general,
these observations suggest that the absolute size of the total NEFA fraction,
rather than its specific composition (aside from EPA), likely influences the 
pathogenesis of diabetes, at least within a population at-risk for diabetes.
